---
title: "Class 10:Structural Bioinformatics"
author: "Rogelio Castro"
format: html
editor: visual
---

# Comparative analysis of ADK
ADK (Adenelate Kinase) is an important drug tarfet and we would love to know how it works- i.e. molecular mechanism.

There has been lots of work done on this protein due to it's importance including lots of crystal structures.

We will begin with getting an example ADK sequence from the database.
We will then use this to find all ADK structures in the PDB.

```{r}
library(bio3d)

aa <- get.seq("1AKE_A")
aa
```

We can now run BLAST with this sequence:

```{r}
b <- blast.pdb(aa)
```

```{r}
hits <- plot(b)
```

Let's see what is in our `hits` object:

```{r}
#hits$pdb.id
```

Now we can download all these PDB structures files:

```{r}
# Download releated PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

Now I want to allign and supperpose these structures which are all over the place.

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```


Let's have a look at our `pdbs` object:

```{r}
pdbs
```

Now we have our aligned and superposed structures we can perform all sorts of analysis on them. Let's do PCA.


```{r}
# Perform PCA
pc.xray <- pca(pdbs)
plot(pc.xray)
```

We can luster the structures by RMSD

```{r}
rd <- rmsd(pdbs)

# Structure-based clustering

hc.rd<- hclust(dist(rd))
grps.rd<- cutree(hc.rd, k=3)

plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```



We can make a movie - also called trajectory of the major differences (i.e. structural displacements) of ADK.

```{r}
# Visualize first principal component
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```

>Q10. Which of the packages above is found only on BioConductor and not CRAN? 

devtools::install_bitbucket("Grantlab/bio3d-view")

>Q11. Which of the above packages is not found on BioConductor or CRAN?: 

bio3d

>Q12. True or False? Functions from the devtools package can be used to install packages from GitHub and BitBucket? 

True

>Q13. How many amino acids are in this sequence, i.e. how long is this sequence? 

214


